Simple repression draft

### 2020-10-16 doc/resubmission/section_02_means.tex
minimal changes to fix some errors

\subsubsection{Two- and Three-state Thermodynamic Models}
We begin our analysis with models 1 and 2 in
Figure~\ref{fig1:means_cartoons}(B). In each of these models the promoter
is idealized as existing in a set of discrete states; the difference being
whether or not the RNAP bound state is included or not. Gene expression is then
assumed to be proportional to the probability of the promoter being in either
the empty state (model 1) or the RNAP-bound state (model (2)). We direct the
reader to Appendix~\ref{sec:non_bursty} for details on the derivation of the
fold-change. For our purposes here, it suffices to state that the functional
form of the fold-change for model 1 is
\begin{equation}
\text{fold-change}
= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
where $R$ is the number of repressors per cell, $N_{NS}$ is the number of
non-specific binding sites where the repressor can bind, $\Delta\varepsilon_R$
is the repressor-operator binding energy, and $\beta \equiv (k_BT)^{-1}$. This
equation matches the form of the master curve in
Figure~\ref{fig1:means_cartoons}(D) with $\rho=1$ and $\Delta F_R =
\beta\Delta\varepsilon_r - \log (R / N_{NS})$. For model 2 we have a similar
situation. The fold-change takes the form
\begin{align}
\text{fold-change}
&= \left(
1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}}
        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
\right)^{-1}
\\
&= (1 + \exp(-\Delta F_R + \log\rho))^{-1},
\end{align}
where $P$ is the number of RNAP per cell, and $\Delta\varepsilon_P$ is the
RNAP-promoter binding energy. For this model we have $\Delta F_R =
\beta\Delta\varepsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
\frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$. Thus far, we see that
the two thermodynamic models, despite making different coarse-graining
commitments, result in the same functional form for the fold-change in mean gene
expression.  We now explore how kinetic models fare when faced with computing
the same observable.

@@

Again, we consign all details of the derivation to
Appendix~\ref{sec:non_bursty}. Here we just highlight the general findings for
all five kinetic models. As already shown in Figure~\ref{fig1:means_cartoons}(C)
and (D), all the kinetic models explored can be collapsed onto the master curve.
Given that the repressor-bound state only connects to the rest of the promoter
dynamics via its binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively,
all models can effectively be separated into two categories: a single
repressor-bound state, and all other promoter states with different levels of
coarse graining. This structure then guarantees that, at steady-state, detailed
balance between these two groups is satisfied. What this implies is that the
steady-state distribution of each of the non-repressor states has the same
functional form with or without the repressor, allowing us to write the
fold-change as a product of the ratio of the binding and unbinding rates of the
promoter, and the promoter details. This results in a fold-change of the form
\begin{align}
\text{fold-change} &= \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
&= (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
\end{align}
where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$
for each model change as shown in Figure~\ref{fig1:means_cartoons}(C). Another
intuitive way to think about these two terms is as follows: in all kinetic
models shown in Figure~1(C) the repressor-bound state can only be reached from a
single repressor-free state. The ratio of these two states --repressor-bound and
adjacent repressor-free state-- must remain the same for all models, regardless
of the details included in other promoter states if $\Delta F_R$ represents an
effective free energy of the repressor binding the DNA operator. The presence of
other states then draws probability density from the promoter being in either of
these two states, making the ratio between the repressor-bound state and
\textit{all} repressor-free states different. The log difference in this ratio
is given by $\log(\rho)$. Since model 1 and model 5 of Figure~1(C) consist of a
single repressor-free state, $\rho$ is then necessarily 1 (See
Appendix~\ref{sec:non_bursty} for further details).

>>>>

\subsubsection{Two- and Three-state Thermodynamic Models}
We begin our analysis with models 1 and 2 in
Figure~\ref{fig1:means_cartoons}(B). In each of these models the promoter
is idealized as existing in a set of discrete states; the difference being
whether or not the RNAP bound state is included or not. Gene expression is then
assumed to be proportional to the probability of the promoter being in either
the empty state (model 1) or the RNAP-bound state (model (2)). We direct the
reader to Appendix~\ref{sec:non_bursty} for details on the derivation of the
fold-change. For our purposes here, it suffices to state that the functional
form of the fold-change for model 1 is
\begin{equation}
\text{fold-change}
= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
where $R$ is the number of repressors per cell, $N_{NS}$ is the number of
non-specific binding sites where the repressor can bind, $\Delta\varepsilon_R$
is the repressor-operator binding energy, and $\beta \equiv (k_BT)^{-1}$. This
equation matches the form of the master curve in
Figure~\ref{fig1:means_cartoons}(D) with $\rho=1$ and $\Delta F_R =
\beta\Delta\varepsilon_r - \log (R / N_{NS})$. For model 2 we have a similar
situation. The fold-change takes the form
\begin{eqnarray}
\text{fold-change}
&=& \left(
1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}}
        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
\right)^{-1}
\\
&=& (1 + \exp(-\Delta F_R + \log\rho))^{-1},
\end{eqnarray}
where $P$ is the number of RNAP per cell, and $\Delta\varepsilon_P$ is the
RNAP-promoter binding energy. For this model we have $\Delta F_R =
\beta\Delta\varepsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
\frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$. Thus far, we see that
the two thermodynamic models, despite making different coarse-graining
commitments, result in the same functional form for the fold-change in mean gene
expression.  We now explore how kinetic models fare when faced with computing
the same observable.

@@

Again, we consign all details of the derivation to
Appendix~\ref{sec:non_bursty}. Here we just highlight the general findings for
all five kinetic models. As already shown in Figure~\ref{fig1:means_cartoons}(C)
and (D), all the kinetic models explored can be collapsed onto the master curve.
Given that the repressor-bound state only connects to the rest of the promoter
dynamics via its binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively,
all models can effectively be separated into two categories: a single
repressor-bound state, and all other promoter states with different levels of
coarse graining. This structure then guarantees that, at steady-state, detailed
balance between these two groups is satisfied. What this implies is that the
steady-state distribution of each of the non-repressor states has the same
functional form with or without the repressor, allowing us to write the
fold-change as a product of the ratio of the binding and unbinding rates of the
promoter, and the promoter details. This results in a fold-change of the form
\begin{eqnarray}
 \text{fold-change} &=& \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
&=& (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
\end{eqnarray}
where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$
for each model change as shown in Figure~\ref{fig1:means_cartoons}(C). Another
intuitive way to think about these two terms is as follows: in all kinetic
models shown in Figure~1(C) the repressor-bound state can only be reached from a
single repressor-free state. The ratio of these two states --repressor-bound and
adjacent repressor-free state-- must remain the same for all models, regardless
of the details included in other promoter states if $\Delta F_R$ represents an
effective free energy of the repressor binding the DNA operator. The presence of
other states then draws probability density from the promoter being in either of
these two states, making the ratio between the repressor-bound state and
\textit{all} repressor-free states different. The log difference in this ratio
is given by $\log(\rho)$. Since model 1 and model 5 of Figure~1(C) consist of a
single repressor-free state, $\rho$ is then necessarily 1 (See
Appendix~\ref{sec:non_bursty} for further details).

### 2020-10-16 doc/resubmission/section_03_beyond_means.tex
minimal changes to fix some errors

One interpretation~\cite{Chong2014, Sevier2016} for the active and inactive
states is that they represent the promoter's supercoiling state: transitions to
the inactive state are caused by accumulation of positive supercoiling, which
inhibits transcription, and transitions back to ``active'' are caused by gyrase
or other topoisomerases relieving the supercoiling. This is an interesting
possibility because it would mean the timescale for promoter state transitions
is driven by topoisomerase kinetics, not by RNAP kinetics. From in vitro
measurements, the former are suggested to be of order minutes~\cite{Chong2014}.
Contrast this with model 2, where the state transitions are assumed to be
governed by RNAP, which, assuming a copy number per cell of order $10^3$, has a
diffusion-limited association rate $k_{on} \sim 10^2~\text{s}^{-1}$ to a target
promoter. Combined with known $K_d$'s of order $\mu$M, this gives an RNAP
dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$. As we will show below,
however, there are some lingering puzzles with interpreting this supercoiling
hypothesis, so we leave it as a speculative hypothesis and refrain from
assigning definite physical meaning to the two states in this model.

Intuitively one might expect that, since transcripts are produced as a Poisson
process only when the promoter is in one of the two states in this model,
transcription initiations should now be ``bunched'' in time, in contrast to the
``anti-bunching'' of models 2 and 3 above. One might further guess that this
bunching would lead to super-Poissonian noise in the mRNA distribution over a
population of cells.  Indeed, as shown in Appendix~\ref{sec:non_bursty}, a
calculation of the Fano factor produces
\begin{align}
\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
\label{eq:model4_fano}
\end{align}
which is strictly greater than 1, verifying the above intuition. Note we have
dropped the $P$ label on the promoter switching rates to emphasize that these
very likely do not represent kinetics of RNAP itself. This calculation can also
be sidestepped by noting that the model is mathematically equivalent to the
simple repression model from~\cite{Jones2014}, with states and rates relabeled
and reinterpreted.

@@

Let us now verify this intuition with a precise derivation to check that $r/k^-$
is in fact the mean burst size and to obtain the full burst size distribution.
Consider first a constant, known dwell time $T$ in the active state. Transcripts
are produced at a rate $r$ per unit time, so the number of transcripts $n$
produced during $T$ is Poisson distributed with mean $rT$, i.e.,
\begin{equation}
    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
\end{equation}
Since the dwell time $T$ is unobservable from single molecule mRNA counts, we
actually want $P(n)$, the dwell time distribution with no conditioning on $T$.
Basic rules of probability theory tell us we can write $P(n)$ in terms of
$P(n\mid T)$ as
\begin{equation}
    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
\end{equation}
But we know the dwell time distribution $P(T)$, which is exponentially
distributed according to
\begin{equation}
    P(T) = k^- \exp(-T k^-),
\end{equation}
so $P(n)$ can be written as
\begin{equation}
    P(n) = k^- \frac{r^n}{n!}
            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
\end{equation}
A standard integral table shows $\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
\begin{equation}
    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
        = \frac{k^-}{k^- + r}
            \left(\frac{r}{k^- + r}\right)^n
        = \frac{k^-}{k^- + r}
            \left(1 - \frac{k^-}{k^- + r}\right)^n,
\end{equation}
which is exactly the geometric distribution with standard parameter
$\theta\equiv k^-/(k^- + r)$ and domain $n \in \{0, 1, 2, \dots\}$.
The mean of the geometric distribution, with this convention, is
\begin{align}
\langle n\rangle = \frac{1 - \theta}{\theta}
        = \left(1 - \frac{k^-}{k^- + r}\right)
                    \frac{k^- + r}{k^-}
        = \frac{r}{k^-},
\end{align}
exactly as we guessed intuitively above.

So in taking the limit $r,k^-\rightarrow\infty$, $r/k^-\equiv b$, we obtain a
model which effectively has only a single promoter state, which initiates bursts
at rate $k^+$ (transitions to the active state, in the model 4 picture). The
master equation for mRNA copy number $m$ as derived in
Appendix~\ref{sec:non_bursty} takes the form
\begin{align}
\begin{split}
{d \over dt}p(m,t) = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
        &+ \sum_{m'=0}^{m-1} k_i p(m',t) \text{Geom}(m-m';b)
         - \sum_{m'=m+1}^\infty k_i p(m,t) \text{Geom}(m'-m;b),
\end{split}
\end{align}
where we use $k_i$ to denote the burst initiation rate, $\text{Geom}(n;b)$ is
the geometric distribution with mean~$b$, i.e., $\text{Geom}(n;b) =
\frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$ (with domain over nonnegative
integers as above). The first two terms are the usual mRNA degradation terms.
The third term enumerates all ways the system can produce a burst of transcripts
and arrive at copy number $m$, given that it had copy number $m'$ before the
burst. The fourth term allows the system to start with copy number $m$, then
produce a burst and end with copy number $m'$. In fact this last sum has trivial
$m'$ dependence and simply enforces normalization of the geometric distribution.
Carrying it out we have
\begin{equation}
\begin{split}
{d \over dt} p(m,t) = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
        &+ \sum_{m'=0}^{m-1} k_i p(m',t) \text{Geom}(m-m';b)
            - k_i p(m,t),
\end{split}
\end{equation}
We direct readers again to Appendix~\ref{sec:gen_fcn_appdx} for further details.
This improves on model 4 in that now the parameters are easily inferred, as we
will see later, and have clean interpretations. The non-Poissonian features are
attributed to the empirically well-established phenomenological picture of
bursty transcription.

>>>>

One interpretation~\cite{Chong2014, Sevier2016, Sevier2018} for the active and
inactive states is that they represent the promoter's supercoiling state:
transitions to the inactive state are caused by accumulation of positive
supercoiling, which inhibits transcription, and transitions back to ``active''
are caused by gyrase or other topoisomerases relieving the supercoiling. This is
an interesting possibility because it would mean the timescale for promoter
state transitions is driven by topoisomerase kinetics, not by RNAP kinetics.
From in vitro measurements, the former are suggested to be of order
minutes~\cite{Chong2014}. Contrast this with model 2, where the state
transitions are assumed to be governed by RNAP, which, assuming a copy number
per cell of order $10^3$, has a diffusion-limited association rate $k_{on} \sim
10^2~\text{s}^{-1}$ to a target promoter. Combined with known $K_d$'s of order
$\mu$M, this gives an RNAP dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$.
As we will show below, however, there are some lingering puzzles with
interpreting this supercoiling hypothesis, so we leave it as a speculative
hypothesis and refrain from assigning definite physical meaning to the two
states in this model.

Intuitively one might expect that, since transcripts are produced as a Poisson
process only when the promoter is in one of the two states in this model,
transcription initiations should now be ``bunched'' in time, in contrast to the
``anti-bunching'' of models 2 and 3 above. One might further guess that this
bunching would lead to super-Poissonian noise in the mRNA distribution over a
population of cells.  Indeed, as shown in Appendix~\ref{sec:non_bursty}, a
calculation of the Fano factor produces
\begin{equation}
\nu = 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
\label{eq:model4_fano}
\end{equation}
which is strictly greater than 1, verifying the above intuition. Note we have
dropped the $P$ label on the promoter switching rates to emphasize that these
very likely do not represent kinetics of RNAP itself. This calculation can also
be sidestepped by noting that the model is mathematically equivalent to the
simple repression model from~\cite{Jones2014}, with states and rates relabeled
and reinterpreted.

@@

Let us now verify this intuition with a precise derivation to check that $r/k^-$
is in fact the mean burst size and to obtain the full burst size distribution.
Consider first a constant, known dwell time $T$ in the active state. Transcripts
are produced at a rate $r$ per unit time, so the number of transcripts $n$
produced during $T$ is Poisson distributed with mean $rT$, i.e.,
\begin{equation}
    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
\end{equation}
Since the dwell time $T$ is unobservable from single molecule mRNA counts, we
actually want $P(n)$, the dwell time distribution with no conditioning on $T$.
Basic rules of probability theory tell us we can write $P(n)$ in terms of
$P(n\mid T)$ as
\begin{equation}
    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
\end{equation}
But we know the dwell time distribution $P(T)$, which is exponentially
distributed according to
\begin{equation}
    P(T) = k^- \exp(-T k^-),
\end{equation}
so $P(n)$ can be written as
\begin{equation}
    P(n) = k^- \frac{r^n}{n!}
            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
\end{equation}
A standard integral table shows $\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
\begin{equation}
    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
        = \frac{k^-}{k^- + r}
            \left(\frac{r}{k^- + r}\right)^n
        = \frac{k^-}{k^- + r}
            \left(1 - \frac{k^-}{k^- + r}\right)^n,
\end{equation}
which is exactly the geometric distribution with standard parameter
$\theta\equiv k^-/(k^- + r)$ and domain $n \in \{0, 1, 2, \dots\}$.
The mean of the geometric distribution, with this convention, is
\begin{equation}
\langle n\rangle = \frac{1 - \theta}{\theta}
        = \left(1 - \frac{k^-}{k^- + r}\right)
                    \frac{k^- + r}{k^-}
        = \frac{r}{k^-},
\end{equation}
exactly as we guessed intuitively above.

So in taking the limit $r,k^-\rightarrow\infty$, $r/k^-\equiv b$, we obtain a
model which effectively has only a single promoter state, which initiates bursts
at rate $k^+$ (transitions to the active state, in the model 4 picture). The
master equation for mRNA copy number $m$ as derived in
Appendix~\ref{sec:non_bursty} takes the form
\begin{equation}
\begin{split}
{d \over dt}p(m,t) = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
        &+ \sum_{m'=0}^{m-1} k_i p(m',t) \text{Geom}(m-m';b)
         - \sum_{m'=m+1}^\infty k_i p(m,t) \text{Geom}(m'-m;b),
\end{split}
\end{equation}
where we use $k_i$ to denote the burst initiation rate, $\text{Geom}(n;b)$ is
the geometric distribution with mean~$b$, i.e., $\text{Geom}(n;b) =
\frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$ (with domain over nonnegative
integers as above). The first two terms are the usual mRNA degradation terms.
The third term enumerates all ways the system can produce a burst of transcripts
and arrive at copy number $m$, given that it had copy number $m'$ before the
burst. The fourth term allows the system to start with copy number $m$, then
produce a burst and end with copy number $m'$. In fact this last sum has trivial
$m'$ dependence and simply enforces normalization of the geometric distribution.
Carrying it out we have
\begin{equation}
\begin{split}
{d \over dt} p(m,t) = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
        &+ \sum_{m'=0}^{m-1} k_i p(m',t) \text{Geom}(m-m';b)
            - k_i p(m,t),
\end{split}
\end{equation}
We direct readers again to Appendix~\ref{sec:gen_fcn_appdx} for further details.
This improves on model 4 in that now the parameters are easily inferred, as we
will see later, and have clean interpretations. The non-Poissonian features are
attributed to the empirically well-established phenomenological picture of
bursty transcription.
